Recombination shapes the diversification of the wtf meiotic drivers

Version 1

Abstract

Abstract

Meiotic drivers are selfish genetic elements that distort fair segregation. The wtf genes are poison-antidote meiotic drivers that are experiencing rapid diversification in fission yeasts. However, gene duplication alone is insufficient to drive the diversification of wtf genes, given the poison encoded by a newly duplicated wtf gene can be detoxified by the antidote encoded by the original wtf gene. Here, we analyze the evolution of wtf genes across 21 strains of Schizosaccharomyces pombe. Knocking out each of 25 wtf genes in S. pombe strain 972h- separately does not attenuate the yeast growth, indicating that the wtf genes might be largely neutral to their carriers in asexual life cycle. Interestingly, wtf genes underwent recurrent and intricate recombination. As proof-of-principle, we generate a novel meiotic driver through artificial recombination between wtf drivers, and its encoded poison cannot be detoxified by the antidotes encoded by their parental wtf genes but can be detoxified by its own antidote. Therefore, we propose that recombination can generate new meiotic drivers and thus shape the diversification of the wtf drivers.

Introduction

During meiosis, the two alleles at a gene locus are separated into gametes, and each gamete has an equal chance of receiving either allele. This fundamental principle of inheritance, known as Mendel’s law of segregation (Abbott and Fairbanks, 2016), holds across most genetic loci in most sexual species. However, meiotic drivers, a class of selfish genetic elements, subvert fair segregation during gametogenesis and are transmitted to more than one-half (even to all) of the functional gametes produced by a heterozygote (Sandler and Novitski, 1957; Lyttle, 1991; Hurst and Werren, 2002; Bravo Nunez et al., 2018b). Meiotic drivers can spread in a population even when they impose fitness costs on their hosts (Crow, 1991; Lindholm et al., 2016). However, the spread of a meiotic driver can be thwarted by the costs imposed on its carriers or by its genetic suppressors (Lindholm et al., 2016).

The fission yeast wtf (with Tf Long Terminal Repeats) gene family provides an excellent model to study how meiotic drivers act and evolve (Hu et al., 2017; Nuckolls et al., 2017). Many wtf genes are autonomous one-gene poison-antidote meiotic drivers that encode both a spore-killing poison (short isoform) and an antidote to the poison (long isoform) using alternative transcriptional initiation (Hu et al., 2017; Nuckolls et al., 2017; Nuckolls et al., 2022). To archieve meiotic drive, all spores are exposed to the poison, whereas only those that inherit wtf express the antidote and are rescued (Hu et al., 2017; Nuckolls et al., 2017; Nuckolls et al., 2022). Some other wtf genes can act as drive suppressors (Bravo Nunez et al., 2018a; Bravo Nunez et al., 2020a). The poison and the antidote differ only in their N-terminal cytosolic tails containing PY (Leu/Pro-Pro-X-Tyr) motifs. PY motif-dependent ubiquitination promotes the transport of the antidote and the poison (physically interacted with the antidote) from the trans-Golgi network to the endosome, thereby preventing toxicity (Zheng et al., 2023).

The wtf gene family is experiencing rapid diversification: the Schizosaccharomyces pombe reference genome encodes 25 wtf genes, some of which are pseudogenes. The copy numbers of wtf genes vary greatly among different S. pombe strains, and frequent nonallelic gene conversion occurs between wtf genes (Hu et al., 2017; Nuckolls et al., 2017; Eickbush et al., 2019). However, these finding are based on a limited number of strains, and the pattern and extent of recombination in the wtf genes remain to be fully explored. Moreover, wtf driver genes are present in the last common ancestor (LCA) of the fission yeasts S. pombe, S. octosporus, S. osmophilus, and S. cryophilus, indicating that wtf genes have likely maintained the capacity to drive for more than 100 million years (De Carvalho et al., 2022). These fission yeast species carry varying numbers of wtf genes, ranging from 5 to 83 (De Carvalho et al., 2022). Yet, it remains perplexing how wtf genes achieved such diversification. On one hand, gene duplication can give birth to new wtf gene copies. On the other hand, a newly duplicated wtf gene might not drive because the poison produced by the newly duplicated wtf gene can be detoxified by the original wtf gene. Like newly duplicated genes, the most probable fate of a new wtf duplicate is pseudogenization (Lynch, 2007; Innan and Kondrashov, 2010). Thus, the vast majority of new wtf duplicates experience an early exit from the population, most probably never reaching fixation (Lynch, 2007; Innan and Kondrashov, 2010). To become a new driver, the new wtf copy should evolve coupling new poison and new antidote to the new poison through mutations. But the fate-changing mutations are likely to be rare. It follows that gene duplication might be insufficient to drive the diversification of wtf genes.

In this study, we analyzed the diversity and evolution of wtf genes in the genomes of 21 strains of S. pombe that were sequenced using long-read sequencing approaches (Tusso et al., 2022). Through knocking out each of 25 wtf genes in S. pombe laboratory strain 972h-, no significant attenuated growth was observed, indicating wtf genes might be not deleterious in the asexual life cycle. We found that recurrent recombination occurred among wtf genes. We generated a novel meiotic driver through artificial recombination between wtf drivers, and its encoded poison cannot be detoxified by the antidotes encoded by their parental wtf genes but can be detoxified by its own antidote. Therefore, we propose that recombination can generate wtf driver with new poisons and might shape the diversification of wtf genes.

Results

Diversity and evolution of wtf genes in fission yeasts

First, we analyzed the diversity and distribution of wtf genes in fission yeasts. The S. pombe reference genome (strain 972h-) encodes a total of 25 wtf genes. For these 25 wtf genes, the number of exons varies from 3 to 6 (Figure 1A) (Bowen et al., 2003). To investigate the relationship among exons from different wtf genes, we grouped these wtf exons into clusters based on nucleotide identity of 0.50 (Figure 1B and C). The wtf exons were grouped into 10 clusters with >2 members, and 12 exons exist as singletons in the similarity network (Figure 1B and C; Supplementary file 1c). Only exon 1 of all the 25 wtf genes group together in a cluster, indicating that the first exons are well conserved among the wtf genes. No other exon is conserved among all the 25 wtf genes. Therefore, the evolution of the wtf gene structures appear to be highly dynamic.

The diversity and evolution of wtf genes in fission yeasts.

(A) The gene structures of the wtf genes in the S. pombe reference genome. Rectangles and lines represent exons and introns, respectively. Rectangles with the same color indicate exons that share >50% sequence identity. (B) The similarity network of exons of the wtf genes in the S. pombe reference genome. Sequences that share an identity of >50% form a cluster. The colors of wtf exons correspond to these in panel (A). The exons for each cluster were listed in the Supplementary file 1c. (C) The heatmap of nucleotide identity among exons of the wtf genes in S. pombe reference genome. (D) The distribution wtf genes in 21 S. pombe strains. wtf genes were present in 25 genetic loci. Filled circles and empty circles represent the presence or absence of wtf genes, respectively. Red pentagons indicate putative wtf pseudogenes. The size of circles in purple indicates the number of wtf genes. The relationship among 21 S. pombe strains was inferred based on phylogenetic analysis of 30 randomly selected genes. (E) Phylogenetic relationship of wtf genes in 21 S. pombe strains. The phylogenetic tree is reconstructed using the maximum likelihood method. Filled circles in blue indicate wtf genes in S. pombe reference genome. The number of exons, the presence of solo-LTRs, the putative pseudogene status, the presence of intron-1-ATG codon, the presence of exon-2-ATG codon is shown near the corresponding wtf gene. wtf genes from other fission yeast species were collapsed into a triangle.

We next identified wtf genes in 21 strains of S. pombe that were sequenced using long-read sequencing approaches (Supplementary file 1b ) (Tusso et al., 2022). The copy number of wtf genes varies among different S. pombe strains, ranging from 24 (strain JB879) to 37 (strain JB1206) (Figure 1D). Synteny analyses show that the wtf genes are present in 20 genetic loci (Figure 1D). Multiple wtf genes were present in 13 wtf loci. Within 20 wtf loci, at least one wtf gene is present in all of or nearly all of the 21 S. pombe strains, suggesting that these 20 wtf loci might have originated before the LCA of the 21 S. pombe strains. Putative wtf pseudogenes are prevalent in many wtf loci among the 21 S. pombe strains, indicating frequent pseudogenization occurred in the wtf genes. These results indicate that wtf copy number variation is prevalent among S. pombe strains.

We performed phylogenetic analyses of the wtf genes from 21 S. pombe strains and three other fission yeast species (S. octosporus, S. cryophilus, and S. osmophilus) (Figure 1E; Supplementary file 1d). The wtf genes of S. pombe form a monophyletic group. Orthologs of wtf14, wtf7, wtf11, and wtf15 form monophyletic groups, whereas orthologs of other wtf genes show complex phylogenetic mixing, indicating complex recombination might have occurred among these wtf genes (Figure 1E) (Eickbush et al., 2019). Moreover, the wtf genes with 6 exons (including the known functional drivers wtf4 and wtf23) cluster together, and exhibit a ladder-like phylogeny, which might be generated by continual selection driven by antidotes (like the ladder-like phylogeny of influenza A viruses H1N1 and H3N2, which is shaped by continual immune selection (Grenfell et al., 2004; Bedford et al., 2011)). Based on phylogenetic relationship, we divided the wtf genes of 21 S. pombe strains into eight groups, namely groups 1 to 8, among which groups 5 to 8 include orthologs of wtf14, wtf7, wtf11, and wtf15, respectively (Figure 1E). Exon 2 ATG codons (exon-2-ATG) and in-frame ATG within intron 1 and near the start of exon 2 (intron-1-ATG) of wtf genes can encode the start of poison protein isoforms (Hu et al., 2017). We found that most of exon-2-ATG and intron-1-ATG are present within group 1 wtf genes (Figure 1E). A majority of the wtf genes are flanked by solo-LTRs (Bowen et al., 2003) (Figure 1E). However, the solo-LTRs flanking the wtf genes do not cluster together but form many distinct groups, suggesting that solo-LTRs were inserted nearby the wtf genes multiple times (Supplementary file 1a). Together, our results reveal the rapid diversification and turnover of wtf genes in a single fission yeast species.

No attenuated growth of fission yeast without wtf genes

To explore the effect of wtf genes on the fitness of fission yeast, we knocked out each of the 25 wtf genes in the S. pombe laboratory strain 972h-using a method based on homologous recombination (Figure 2A). A total of 25 wtf knockout strains (Δwtf1 to Δwtf25) were generated. We used spot assay to evaluate the effect of wtf gene knockout on the yeast growth, and no growth defect was observed for all the 25 wtf knockout strains (Figure 2B). Therefore, our experiment suggests that the wtf genes might be largely neutral to the fitness of their carriers in the asexual life cycle at least in normal growth condition.

The effect of wtf gene on the growth of S. pombe.

(A) Generation of wtf knockout (Δwtf) strains based on the homologous recombination method. Substitution cassette contains a kanMX resistance marker and two homologous sequences flanking the target wtf genes. (B) Spot assay of Δwtf strains. The strains were continuously diluted by a 10-fold gradient to 10-5, and five stock suspensions were spotted onto the YE (Yeast extract) liquid medium.

Recurrent recombination in wtf genes

Given complex phylogenetic mixing observed among wtf genes (Figure 1E), we tested whether recombination occurred. We detected signals of recombination in the 25 wtf genes of the S. pombe reference genome (p = 0) and in the wtf genes of the 21 S. pombe strains (p = 0) using pairwise homoplasy index (HPI) test. Split network analysis also supports the frequent occurrence of recombination in the 25 wtf genes of the S. pombe reference genome (Figure 3A) and in the wtf genes of the 21 strains of S. pombe (Figure 3B). In contrast, no recombination signal was tested for groups 5 to 8 using HPI test (p = 1 for group 5, p = 1 for group 6, p = 0.53 for group 7, and p = 1 for group 8). We estimated recombination rates of the full-length wtf sequences, the first exons, and the wtf sequences without the first exons for wtf groups 1 to 4. We found that the recombination rate of group 1 wtf was highest among the four wtf groups (Figure 3C). For group 1, breakpoints are dispersed across the wtf sequences (Figure 3D). These lines of evidence suggest that wtf genes underwent recurrent and intricate recombination.

Recombination analysis of wtf genes.

(A) Split tree of 25 wtf genes of the S. pombe reference genome. (B) Split tree of wtf genes from 21 S. pombe strains. The wtf genes of the S. pombe reference genome are labelled. (C) Recombination rates of wtf genes that belong to groups 1 to 4. Recombination rates were estimated for the full-length wtf sequences, the first exons, and the wtf sequences without the first exons. (D) Breakpoints detected for wtf genes of group1. The exons of wtf4, as a gene position reference, are shown.

Generation of a new driver gene through artificial recombination

Given gene duplication alone might be insufficient to shape the diversification of wtf genes, we hypothesize that recombination between wtf genes can generate new meiotic drives. To test this, we constructed four chimeric wtf genes through recombination among known functional meiotic drivers (wtf23) and an artificially generated meiotic driver (wtf18).

We used a proved Saccharomyces cerevisiae system to test the activity of poison and antidote proteins encoded by wtf genes (Nuckolls et al., 2020). As expected, the expression of the poison proteins (Wtf23poison) encoded by wtf23 genes caused the yeast growth arrest (Figure 4A). The attenuated growth was alleviated, when the corresponding antidote proteins (Wtf23antidote) were expressed (Figure 4A). We also experimentally analyzed wtf18 gene, which was known to encode only long (antidote-like) transcripts and probably act as a suppressor (Bravo Nunez et al., 2018a). We artificially introduced an in-frame ATG codon right before the start of exon 2, generating wtf18poison/-0M. The expression of wtf18poison/-0Mresulted in the yeast growth arrest, suggesting its product, Wtf18poison/-0M, is indeed a poison protein (Figure 4B). When co-expressing wtf18antidote and wtf18poison/-0M, the attenuated yeast growth was rescued (Figure 4B), indicating that Wtf18antidote can ameliorate the toxicity of Wtf18poison/-0M. Therefore, the introduction of an ATG codon to wtf18 leads to the generation of a putative meiotic driver.

Poison and antidote activity of wtf genes and chimeric wtf genes.

wtf23 and wtf18 are highlighted in green and red, respectively. Rectangles represent exons. The start codon ATG is shown. For wtf18, an in-frame ATG codon was introduced right before the start of exon 2, generating wtf18poison/-0M. Spot assay of yeast transformed with the short isoforms (encoding poison-like proteins) and the long isoforms (encoding antidote-like proteins) of wtf23 (A) and wtf18 (B). (C to F) Four chimeric wtf genes were generated through artificial recombination, namely wtfC1 (with exons 1-2 of wtf23 and exons 3-6 of wtf18), wtfC2 (with exons 1-3 of wtf23 and exons 4-6 of wtf18), wtfC3 (with exons 1-4 of wtf23 and exons 5-6 of wtf18), and wtfC4 (with exons 1-5 of wtf23 and exon 6 of wtf18). Spot assay of yeast transformed with the short isoforms (encoding poison-like proteins) and the long isoforms (encoding antidote-like proteins) of wtfC1 (C), wtfC2 (D), wtfC3 (E), and wtfC4 (F) are shown.

We then constructed four chimeric wtf genes through artificial recombination between wtf23 and wtf18, including wtfC1 (possessing exons 1-2 of wtf23 and exons 3-6 of wtf18), wtfC2 (possessing exons 1-3 of wtf23 and exons 4-6 of wtf18), wtfC3 (possessing exons 1-4 of wtf23 and exons 5-6 of wtf18), and wtfC4 (possessing exons 1-5 of wtf23 and exon 6 of wtf18). The expression of the short isoforms of wtfC1, wtfC2, wtfC3, and wtfC4 resulted in yeast growth arrest, revealing their toxicity (Figure 4C-F). However, the antidote of wtfC1 and wtfC3 cannot detoxify the corresponding chimeric toxins (Figure 4C and E). Interestingly, we generated a putative novel meiotic driver, namely wtfC4. Our results show that wtfC4 encodes a functional poison (WtfC4poison) (Figure 4F). The poison can be detoxified by its own long isoforms (dubbed as wtfC4antidote), but cannot be detoxified by the antidote proteins of their parental genes (Figure 4F). Taken together, we generated a new meiotic driver through artificial recombination between pre-existing wtf genes.

Discussion

In this study, we analyzed the diversity and evolution of wtf genes in fission yeasts. The copy number of the wtf gene varies among different S. pombe strains, revealing rapid diversification and turnover of the wtf genes within a single fission yeast species. We detected signals of recurrent and intricate recombination among wtf genes as previously reports with limited genomes (Hu et al., 2017; Nuckolls et al., 2017; Eickbush et al., 2019; De Carvalho et al., 2022). We hypothesize that recombination between wtf genes can produce new wtf genes with new poisons and the antidotes to new poisons. These new wtf genes can then driver through populations. As proof-of-principle, we generated a chimeric wtf gene that represents a new meiotic driver. The encoded poison of the newly generated meiotic driver can be detoxified by its own long isoforms, but cannot be detoxified by the antidote proteins of their parental genes. However, the other three chimeric wtf genes did not show this property, possibly because the last exon might be crucial for antidote function. Indeed, our recombination breakpoint analyses (Figure 3D) reveal substantial recombination might have occurred in the last exon. Together, our results indicate that recombination is likely to drive the rapid diversification of wtf gene in fission yeasts.

Most of the known meiotic drivers impose costs on their carriers due to direct effects of the driver on survival or fertility, production of a biased sex ratio, or via deleterious mutations linked to the driver (Price and Wedell, 2008; Larracuente et al., 2012; Sutter and Lindholm, 2015; Fishman et al., 2015; Lindholm et al., 2016; Zanders and Unckless, 2019). In outcrossing between individuals from distinct yeast lineages, wtf drivers can provide a selective advantage to atypical spores, such as aneuploids and diploids (Bravo Nunez et al., 2020b). In this study, we assessed the effects of the wtf genes on the growth of fission yeast during the asexual life cycle through knocking out each of the 25 wtf genes in S. pombe laboratory strain 972h-separately. We did not observe obvious attenuated growth for these wtf knockout strains, indicating wtf genes are largely neutral to the fitness of their carriers during the asexual life cycle at least in the normal growth setting. It should be noted that the spot assay used in this study detect large differences in fitness between wide type and wtf knockout strains. Nevertheless, it is likely that wtf genes evolve mainly in a neutral manner during the asexual life cycle, which explains the presence of high proportion of pseudogenes in wtf gene repertoire. Moreover, asexual reproduction is much more frequent than sexual reproduction for yeasts (Tsai et al., 2008). Therefore, even if fate-changing mutations that simultaneously produce new poison and the antidote to new poison occur, the most probable fate of a new wtf gene generated by gene duplication is pseudogenization and removed from the population.

Gene duplication gives rise to new wtf genes. However, the newly generated wtf gene can be detoxified by the original wtf gene and thus cannot drive through its host population, when the original wtf is fixed in the population (Figure 5). Therefore, most, if not all, of wtf gene duplicates experience early exist from the host population. When recombination occurs between two pre-existing wtf genes, chimeric wtf gene with new poison and the antidote to new poison can be generated as this study shows. Then, the wtf gene with new driver property can spread in its host population, even reaching fixation (Figure 5). During asexual life cycle, wtf genes evolve mainly under genetic drift, and thus can accumulate disruptive mutations, leading to their pseudogenization. Taken together, our study highlights the significance of recombination in shaping the diversification of wtf genes.

The evolutionary mechanisms of wtf genes.

(A) Mechanism of wtf meiotic driver. wtf genes are transcribed into two transcripts, namely long and short isoforms. The long and short isoforms encode antidote and poison, respectively. All spores are exposed to the poison, whereas only those that inherit wtf genes express the antidote and are rescued. Spores without the wtf allele are destroyed, resulting in the drive of wtf. (B) Duplication is insufficient to drive the diversification of wtf genes. Gene duplication of wtfA gives rise to a new wtf gene, wtfAdup. However, wtfAdupgene can be detoxified by the original wtfA gene and thus cannot drive through its host population. (C) When recombination occurs between two parental wtfB and wtfC genes, generating chimeric wtfD gene. wtfD encodes new poison that cannot be detoxified by their parental antidote and the antidote to new poison. The wtfD gene then can spread in its host population.

Methods

Identification of the wtf genes

We used the blastn algorithm to identify wtf genes within 21 S. pombe strains with 25 wtf genes from S. pombe reference genome as the queries and an e-cutoff value of 10-5. The identified wtf genes were annotated based on the wtf genes of reference genome. The sequence identity among the exons of wtf genes was calculated using BioAider version 1.334 (Zhou et al., 2020). Exons was then clustered based on the nucleotide identity using igraph package version 2.0.1.1 (Csardi et al., 2006; Csardi et al., 2024). We extended 1000 bp flanking each wtf gene to establish their syntenic relationships.

Phylogenetic analysis

The coding sequences of wtf genes of 21 S. pombe strains and three other fission yeast species (S. octosporus, S. cryophilus, and S. osmophilus) were aligned using MAFFT version 7 (Katoh and Standley, 2013). To clarify the relationship of 21 S. pombe strains, 30 genes were randomly selected and concatenated using Phylosuite version 1.2.1 (Zhang et al., 2020). All the phylogenetic analyses in this study were performed using the maximum likelihood (ML) method implemented in IQ-TREE version 2 (Minh et al., 2020). The best-fit substitution model was selected using the ModelFinder algorithm (Kalyaanamoorthy et al., 2017). Node supports were assessed using the ultrafast bootstrap approximation (UFBoot) method with 1,000 replicates (Hoang et al., 2018). Solo-LTRs were identified using the blast algorithm and aligned using MAFFT version 7 (Katoh and Standley, 2013). Phylogenetic analysis was performed using the approximate maximum likelihood method implemented in FastTree version 2.1.1 (Price et al., 2010).

Recombination analysis

Split networks of wtf genes were generated using the neighborhood network analysis implemented in Splittree4 (Huson et al., 2006). Pairwise homoplasy index (PHI) test was performed using SplitsTree4 (Huson et al., 2006). Potential breakpoints were detected using 3SEQ (Lam et al., 2018). The recombination rate was estimated using FastEPRR package version 2.0 (Gao et al., 2016).

Generation of wtf knockout strains

The wtf knockout (Δwtf) strains generated in this study were derived from S. pombe strain 972h-. We constructed substitution cassettes for each of the 25 wtf genes of the S. pombe reference genome. Substitution cassettes contain a kanMX resistance marker and two homologous sequences flanking the target wtf genes (Moreno et al., 1991; García-Ríos et al., 2014). Substitution cassettes were transformation into S. pombe (strain 972h-) through the lithium acetate-based method (Moreno et al., 1991; García-Ríos et al., 2014). wtf gene knockout strains were selected for kanMX resistance and were verified by PCR.

Plasmid construction

Total RNA of fission yeast was extracted and reverse transcribed into cDNA. Coding sequences of wtf23antidote, wtf23poison, and wtf18antidote were amplified using the corresponding primers (Supplementary file 1e). wtf18poison/-M0 was generated using wtf18antidote as the template and the primer with an artificially introduced ATG (Supplementary file 1e). We generated wtfC1 through recombining exons 1-2 of wtf23 and exons 3-6 of wtf18, generated wtfC2 through recombining exons 1-3 of wtf23 and exons 4-6 of wtf18, generated wtfC3 through recombining exons 1-4 of wtf23 and exons 5-6 of wtf18, and generated wtfC4 through recombining exons 1-5 of wtf23 and exon 6 of wtf18. These wtf and wtfC genes were then cloned into the GAL1/10 dual expression plasmid Gal_HF. Plasmids were first transformed into Escherichia coli and verified by PCR and sequencing. Plasmids were then transformed into S. cerevisiae (strain S288C) using the lithium acetate-based method. Yeast transformants were selected for kanMX resistance and were verified by PCR.

Spot assay

The yeast strains were cultured in YPD liquid medium at 30℃ with shaking at 200 rpm. The overnight cultures were transferred to fresh YPD liquid medium and grown to an OD600 value of ∼3. Cells were collected by centrifugation, and the OD600 was adjusted to 3. Subsequently, the strains were continuously diluted by a 10-fold gradient to 10-5, generating five stock suspensions. The five stock suspensions were used for spot assays. We plated 1.5μL of each stock suspension on the surface of YPD and YPG solid media. The plates were incubated at 30℃, and the growth of colonies were observed.

eLife Assessment

This important study provides one mechanism that can explain the rapid diversification of poison-antidote pairs in fission yeast: recombination between existing pairs. The evidence is largely solid, but the study can benefit from demonstrating that the novel poison-antidote constructed by the authors can serve as a meiotic driver. The work is of interest to colleagues studying genetic incompatibilities.

Reviewer #1 (Public review):

Summary

The authors determine the phylogenetic relation of the roughly two dozen wtf elements of 21 S. pombe isolates and show that none of them in the original S. pombe are essential for robust mitotic growth. It would be interesting to test their meiotic function by simply crossing each deletion mutant with the parent and analyzing spores for non-Mendelian inheritance. If this has been reported already, that information should be added to the manuscript. If not, I suggest the authors do these simple experiments and add this information.

Strengths:

The most interesting data (Figure 4) show that one recombinant (wtfC4) between wtf18 and wtf23 produces in mitotic growth a poison counteracted by its own antidote but not by the parental antidotes. Again, it would be interesting to test this recombinant in a more natural setting - meiosis between it and each of the parents.

Weaknesses:

In the opinion of this reviewer, some minor rewriting is needed.

Reviewer #2 (Public review):

Summary:

This important study provides a mechanism that can explain the rapid diversification of poison-antidote pairs (wtf genes) in fission yeast: recombination between existing genes.

Strengths:

The authors analyzed the diversity of wtf in S. pombe strains, and found pervasive copy number variations. They further detected signals of recurrent recombination in wtf genes. To address whether recombination can generate novel wtf genes, the authors performed artificial recombination between existing wft genes, and showed that indeed a new wtf can be generated: the poison cannot be detoxified by the antidotes encoded by parental wtf genes but can be detoxified by own antidote.

Weaknesses:

The study can benefit from demonstrating that the novel poison-antidote constructed by the authors can serve as a meiotic driver.

Reviewer #3 (Public review):

Summary:

In this manuscript, Wang and colleagues explore factors contributing to the diversification of wtf meiotic drivers. wtf genes are autonomous, single-gene poison-antidote meiotic drivers that encode both a spore-killing poison (short isoform) and an antidote to the poison (long isoform) through alternative transcriptional initiation. There are dozens of wtf drivers present in the genomes of various yeast species, yet the evolutionary forces driving their diversification remain largely unknown. This manuscript is written in a straightforward and effective manner, and the analyses and experiments are easy to follow and interpret. While I find the research question interesting and the experiments persuasive, they do not provide any deeper mechanistic understanding of this gene family.

Strengths:

(1) The authors present a comprehensive compendium and analysis of the evolutionary relationships among wtf genes across 21 strains of S. pombe.

(2) The authors found that a synthetic chimeric wtf gene, combining exons 1-5 of wtf23 and exon 6 of wtf18, behaves like a meiotic driver that could only be rescued by the chimeric antidote but neither of the parental antidotes. This is a very interesting observation that could account for their inception and diversification.

Weaknesses:

(1) Deletion strains

The authors separately deleted all 25 Wtf genes in the S. pombe ference strain. Next, the authors performed a spot assay to evaluate the effect of wtf gene knockout on the yeast growth. They report no difference to the WT and conclude that the wtf genes might be largely neutral to the fitness of their carriers in the asexual life cycle at least in normal growth conditions.

The authors could have conducted additional quantitative growth assays in yeast, such as growth curves or competition assays, which would have allowed them to detect subtle fitness effects that cannot be quantified with a spot assay. Furthermore, the authors do not rule out simpler explanations, such as genetic redundancy. This could have been addressed by crossing mutants of closely related paralogs or editing multiple wtf genes in the same genetic background.

Another concern is the lack of detailed information about the 25 knockout strains used in the study. There is no information provided on how these strains were generated or, more importantly, validated. Many of these wtf genes have close paralogs and are flanked by repetitive regions, which could complicate the generation of such deletion strains. As currently presented, these results would be difficult to replicate in other labs due to insufficient methodological details

(2) Lack of controls

The authors found that a synthetic chimeric wtf gene, constructed by combining exons 1-5 of wtf23 and exon 6 of wtf18, behaves as a meiotic driver that can be rescued only by its corresponding chimeric antidote, but not by either of the parental antidotes (Figure 4F). In contrast, three other chimeric wtf genes did not display this property (Figure 4C-E). No additional experiments were conducted to explain these differences, and basic control experiments, such as verifying the expression of the chimeric constructs, were not performed to rule out trivial explanations. This should be at the very least discussed. Also, it would have been better to test additional chimeras.

3. Statistical analyses

In line 130 the authors state that: "Given complex phylogenetic mixing observed among wtf genes (Figure 1E), we tested whether recombination occurred. We detected signals of recombination in the 25 wtf genes of the S. pombe reference genome (p = 0) and in the wtf genes of the 21 S. pombe strains (p = 0) using pairwise homoplasy index (HPI) test. ". Reporting a p-value of 0 is not appropriate. Exact P-values should be reported.

Author response:

The following is the authors’ response to the original reviews.

Public Reviews:

Reviewer #1 (Public review):

Summary

The authors determine the phylogenetic relation of the roughly two dozen wtf elements of 21 S. pombe isolates and show that none of them in the original S. pombe are essential for robust mitotic growth. It would be interesting to test their meiotic function by simply crossing each deletion mutant with the parent and analyzing spores for non-Mendelian inheritance. If this has been reported already, that information should be added to the manuscript. If not, I suggest the authors do these simple experiments and add this information.

Thanks for the great summary! All the wtf genes have been tested for meiotic drive phenotypes previously by Bravo Nunez et al. (2020; http://doi.org/10.1371/journal.pgen.1008350). The reference was cited in our original manuscript, and we added the details in the revised manuscript.

Strengths:

The most interesting data (Figure 4) show that one recombinant (wtfC4) between wtf18 and wtf23 produces in mitotic growth a poison counteracted by its own antidote but not by the parental antidotes. Again, it would be interesting to test this recombinant in a more natural setting - meiosis between it and each of the parents.

Thanks for this insightful comment! As suggested, we have tried to test this recombinant in a more natural setting. We created a recombinant strain (wtfC4) based on the laboratory strain 972h-. Specifically, we replaced the last exon of the original wtf23 gene with the last exon of wtf18. However, we encountered a challenge: since strain 972h- has only one mating type and cannot undergo meiosis on its own, we had to mate the recombinant strain with a BN0 h⁺ strain that only carries the wtf23antidote. Unfortunately, despite of tens of attempts over nearly a year, we did not observe meiotic driver phenotype as expected. This might be due to issues with the proper splicing and expression of the potential poison and antidote proteins or due to the genetic background. Similarly, the drive activity of wtf13 has been shown to be specifically suppressed in certain backgrounds.

Weaknesses:

In the opinion of this reviewer, some minor rewriting is needed.

We did the rewriting as this reviewer suggested.

Reviewer #2 (Public review):

Summary:

This important study provides a mechanism that can explain the rapid diversification of poison-antidote pairs (wtf genes) in fission yeast: recombination between existing genes.

Thanks!

Strengths:

The authors analyzed the diversity of wtf in S. pombe strains, and found pervasive copy number variations. They further detected signals of recurrent recombination in wtf genes. To address whether recombination can generate novel wtf genes, the authors performed artificial recombination between existing wft genes, and showed that indeed a new wtf can be generated: the poison cannot be detoxified by the antidotes encoded by parental wtf genes but can be detoxified by own antidote.

Thanks for the great summary!

Weaknesses:

The study can benefit from demonstrating that the novel poison-antidote constructed by the authors can serve as a meiotic driver.

Thanks for this insightful comment! As suggested, we have tried to test this recombinant in a more natural setting. We created a recombinant strain (wtfC4) based on the laboratory strain 972h-. Specifically, we replaced the last exon of the original wtf23 gene with the last exon of wtf18. However, we encountered a challenge: since strain 972h- has only one mating type and cannot undergo meiosis on its own, we had to mate the recombinant strain with a BN0 h⁺ strain that only carries the wtf23antidote. Unfortunately, despite of tens of attempts over nearly a year, we did not observe meiotic driver phenotype as expected. This might be due to issues with the proper splicing and expression of the potential poison and antidote proteins or due to the genetic background. Similarly, the drive activity of wtf13 has been shown to be specifically suppressed in certain backgrounds.

Reviewer #3 (Public review):

Summary:

In this manuscript, Wang and colleagues explore factors contributing to the diversification of wtf meiotic drivers. wtf genes are autonomous, single-gene poison-antidote meiotic drivers that encode both a spore-killing poison (short isoform) and an antidote to the poison (long isoform) through alternative transcriptional initiation. There are dozens of wtf drivers present in the genomes of various yeast species, yet the evolutionary forces driving their diversification remain largely unknown. This manuscript is written in a straightforward and effective manner, and the analyses and experiments are easy to follow and interpret. While I find the research question interesting and the experiments persuasive, they do not provide any deeper mechanistic understanding of this gene family.

Thanks! Please see the following for our point-to-point response.

Strengths:

(1) The authors present a comprehensive compendium and analysis of the evolutionary relationships among wtf genes across 21 strains of S. pombe.

(2) The authors found that a synthetic chimeric wtf gene, combining exons 1-5 of wtf23 and exon 6 of wtf18, behaves like a meiotic driver that could only be rescued by the chimeric antidote but neither of the parental antidotes. This is a very interesting observation that could account for their inception and diversification.

Thanks for the great summary!

Weaknesses:

(1) Deletion strains

The authors separately deleted all 25 Wtf genes in the S. pombe ference strain. Next, the authors performed a spot assay to evaluate the effect of wtf gene knockout on the yeast growth. They report no difference to the WT and conclude that the wtf genes might be largely neutral to the fitness of their carriers in the asexual life cycle at least in normal growth conditions.

The authors could have conducted additional quantitative growth assays in yeast, such as growth curves or competition assays, which would have allowed them to detect subtle fitness effects that cannot be quantified with a spot assay. Furthermore, the authors do not rule out simpler explanations, such as genetic redundancy. This could have been addressed by crossing mutants of closely related paralogs or editing multiple wtf genes in the same genetic background.

Another concern is the lack of detailed information about the 25 knockout strains used in the study. There is no information provided on how these strains were generated or, more importantly, validated. Many of these wtf genes have close paralogs and are flanked by repetitive regions, which could complicate the generation of such deletion strains. As currently presented, these results would be difficult to replicate in other labs due to insufficient methodological details

We generated growth curves for all the 25 wtf deletion strains. We provided the details for wtf gene knockout. However, for 25 wtf genes, there are too many combinations for editing two genes, and it is technically challenging to knock out multiple wtf together. Nevertheless, our results suggest single wtf genes have little effect on the host fitness under normal condition.

(2) Lack of controls

The authors found that a synthetic chimeric wtf gene, constructed by combining exons 1-5 of wtf23 and exon 6 of wtf18, behaves as a meiotic driver that can be rescued only by its corresponding chimeric antidote, but not by either of the parental antidotes (Figure 4F). In contrast, three other chimeric wtf genes did not display this property (Figure 4C-E). No additional experiments were conducted to explain these differences, and basic control experiments, such as verifying the expression of the chimeric constructs, were not performed to rule out trivial explanations. This should be at the very least discussed. Also, it would have been better to test additional chimeras.

We verified the expression of the chimeric genes. The last exon of wtf18 is too small (128bp) to do more meaningful chimeras.

(3) Statistical analyses

In line 130 the authors state that: "Given complex phylogenetic mixing observed among wtf genes (Figure 1E), we tested whether recombination occurred. We detected signals of recombination in the 25 wtf genes of the S. pombe reference genome (p = 0) and in the wtf genes of the 21 S. pombe strains (p = 0) using pairwise homoplasy index (HPI) test." Reporting a p-value of 0 is not appropriate. Exact P-values should be reported.

Due to software limitations, the PHI test reports p-values of 0.0 for extremely significant results. We have therefore reported them as <0.0001 in the revised manuscript.

Recommendations for the authors:

Reviewing Editor Comments:

Regarding the synthetic chimeric wtf gene constructed by combining exons of wtf23 and wtf18, the authors did not explicitly test whether it acts as a meiotic driver in the natural context of a cross. Instead, they examined this possibility only through transgenic overexpression experiments. Given that this is arguably the most important claim of the paper, it is critical that the authors perform, report, and discuss such an experiment in a natural context, regardless of the outcome. It is not necessary to test other recombinants or other wtf loci.

Thanks for this insightful comment! As suggested, we have tried to test this recombinant in a more natural setting. We created a recombinant strain (wtfC4) based on the laboratory strain 972h-. Specifically, we replaced the last exon of the original wtf23 gene with the last exon of wtf18. However, we encountered a challenge: since strain 972h- has only one mating type and cannot undergo meiosis on its own, we had to mate the recombinant strain with a BN0 h⁺ strain that only carries the wtf23antidote. Unfortunately, despite of tens of attempts over nearly a year, we did not observe meiotic driver phenotype as expected. This might be due to issues with the proper splicing and expression of the potential poison and antidote proteins or due to the genetic background. Similarly, the drive activity of wtf13 has been shown to be specifically suppressed in certain backgrounds.

Reviewer #1 (Recommendations for the authors):

The paper is very well written, but some minor points should be corrected or checked.

(1) Line 95: Why "Putative"? Is it not clear what a wtf pseudogene is?

“Putative” was removed.

(2) Line 105: Does "known functional" mean they are active (i.e., have been tested and shown to be active)? If so, a reference should be added.

We used “known meiotic divers”, and added reference here.

(3) Line 135: "no recombination signal was tested". Do the authors mean no signal was inferred?

We changed “tested” to “detected”.

(4) Line 147: References for "known functional meiotic drivers (wtf23) and artificially generated meiotic driver (wtf18)" should be given. A statement of how wtf18 was "artificially generated" is essential so the reader knows how that element differs from the wtfC4 generated here.

Reference for wtf23. As for wtf18, we have specified in the follow text, namely “we artificially introduced an in-frame ATG codon right before the start of exon 2, generating wtf18poison/-0M.”

(5) Lines 154 and 424 say an ATG codon was introduced "right before the start of exon 2," but Figure 4B shows it before exon 1.

We thank the reviewer. The introduced ATG is the second start codon in the long transcript and the first in the short transcript. The right panel of Figure 4B shows the short transcript, so the text and figure are consistent.

(6) Line 159: The wtf18 mutant with this additional ATG codon should be tested in meiosis, to see if "putative" is correct.

Thanks. As wtfC4, we came with technical challenges to show the driver phenotype in a natural setting, and thus removed this statement.

(7) Line 181: change "driver" to "drive".

Driver is correct.

(8) Line 184: insert to read "wtf genes tested". Also, what is the basis for proposing that "the last exon might be crucial for antidote function"?

“Tested” added, and removed the statement.

(9) Line 198: change to read "detects only large differences".

Done as suggested.

(10) Line 204: change "removed" to "removal".

Done as suggested.

(11) Lines 242 and 243: Are "Splittree4" and "SplitsTree4" different, or is this a misprint?

Corrected!

(12) Lines 274-5 and 412 -3 would read better as "strains were diluted in five 10-fold steps” and “...μL of each dilution spotted on” “…to assay for…"

Done as suggested.

(13) Line 284 says "No new data were generated." This is clearly wrong. Perhaps the authors mean there are no supplementary data files.

Corrected!

(14) Line 406: Change "is" to "are".

Corrected!

(15) Line 413: Surely, they were spotted onto YE agar medium, not liquid medium.

Corrected!

(16) Figure 3C: Define "Rho" and the scale used.

The definition of Rho has been added to the Methods section in the revised manuscript.

Reviewer #2 (Recommendations for the authors):

The evidence is largely solid, but the study can benefit from demonstrating that the novel poison-antidote constructed by the authors can serve as a meiotic driver.

As suggested, we have tried to test this recombinant in a more natural setting. We created a recombinant strain (wtfC4) based on the laboratory 972h-. Specifically, we replaced the last exon of the original wtf23 gene with the last exon of wt18f. However, we encountered a challenge: since 972h- is a mating-type strain and cannot undergo meiosis on its own, we had to mate the recombinant strain with a BN0 h⁺ strain that carries the wtf23antidote. Unfortunately, despite of tens of attempts over nearly a year, we did not observe meiotic driver phenotype as expected. This might be due to issues with the proper splicing and expression of the potential poison and antidote proteins.

Reviewer #3 (Recommendations for the authors):

I strongly recommend the authors provide all the details concerning the generation of the knock-out strains, including specific primers used (for both the deletion and validation), the result of these validations, and the specific genotype (and ID) of the strains generated.

These details are now included in the Materials and Methods section and in Supplementary.

Please also provide exact P-values (see point 3).

Due to software limitations, the PHI test reports p-values of 0.0 for extremely significant results. We have therefore reported them as <0.0001 in the revised manuscript.

Version 2

Abstract

Meiotic drivers are selfish genetic elements that distort fair segregation. The wtf genes are poison-antidote meiotic drivers that are experiencing rapid diversification in fission yeasts. However, gene duplication alone is insufficient to drive the diversification of wtf genes, given the poison encoded by a newly duplicated wtf gene can be detoxified by the antidote encoded by the original wtf gene. Here, we analyze the evolution of wtf genes across 21 strains of Schizosaccharomyces pombe. Knocking out each of 25 wtf genes in S. pombe strain 972h-separately does not attenuate the yeast growth, indicating that the wtf genes might be largely neutral to their carriers in asexual life cycle. Interestingly, wtf genes underwent recurrent and intricate recombination. As proof-of-principle, we generate a novel meiotic driver through artificial recombination between wtf drivers, and its encoded poison cannot be detoxified by the antidotes encoded by their parental wtf genes but can be detoxified by its own antidote. Therefore, we propose that recombination can generate new meiotic drivers and thus shape the diversification of the wtf drivers.

Introduction

During meiosis, the two alleles at a gene locus are separated into gametes, and each gamete has an equal chance of receiving either allele. This fundamental principle of inheritance, known as Mendel’s law of segregation (Abbott and Fairbanks, 2016), holds across most genetic loci in most sexual species. However, meiotic drivers, a class of selfish genetic elements, subvert fair segregation during gametogenesis and are transmitted to more than one-half (even to all) of the functional gametes produced by a heterozygote (Sandler and Novitski, 1957; Lyttle, 1991; Hurst and Werren, 2001; Bravo Nunez et al., 2018b). Meiotic drivers can spread in a population even when they impose fitness costs on their hosts (Crow, 1991; Lindholm et al., 2016). However, the spread of a meiotic driver can be thwarted by the costs imposed on its carriers or by its genetic suppressors (Lindholm et al., 2016).

The fission yeast wtf (with Tf Long Terminal Repeats) gene family provides an excellent model to study how meiotic drivers act and evolve (Hu et al., 2017; Nuckolls et al., 2017). Many wtf genes are autonomous one-gene poison-antidote meiotic drivers that encode both a spore-killing poison (short isoform) and an antidote to the poison (long isoform) using alternative transcriptional initiation (Hu et al., 2017; Nuckolls et al., 2017; Nuckolls et al., 2022). To archieve meiotic drive, all spores are exposed to the poison, whereas only those that inherit wtf express the antidote and are rescued (Hu et al., 2017; Nuckolls et al., 2017; Nuckolls et al., 2022). Some other wtf genes can act as drive suppressors (Bravo Nunez et al., 2018a; Bravo Nunez et al., 2020a). The poison and the antidote differ only in their N-terminal cytosolic tails containing PY (Leu/Pro-Pro-X-Tyr) motifs. PY motif-dependent ubiquitination promotes the transport of the antidote and the poison (physically interacted with the antidote) from the trans-Golgi network to the endosome, thereby preventing toxicity (Zheng et al., 2023).

The wtf gene family is experiencing rapid diversification: the Schizosaccharomyces pombe reference genome encodes 25 wtf genes, some of which are pseudogenes. The copy numbers of wtf genes vary greatly among different S. pombe strains, and frequent nonallelic gene conversion occurs between wtf genes (Hu et al., 2017; Nuckolls et al., 2017; Eickbush et al., 2019). However, these finding are based on a limited number of strains, and the pattern and extent of recombination in the wtf genes remain to be fully explored. Moreover, wtf driver genes are present in the last common ancestor (LCA) of the fission yeasts S. pombe, S. octosporus, S. osmophilus, and S. cryophilus, indicating that wtf genes have likely maintained the capacity to drive for more than 100 million years (De Carvalho et al., 2022). These fission yeast species carry varying numbers of wtf genes, ranging from 5 to 83 (De Carvalho et al., 2022). Yet, it remains perplexing how wtf genes achieved such diversification. On one hand, gene duplication can give birth to new wtf gene copies. On the other hand, a newly duplicated wtf gene might not drive because the poison produced by the newly duplicated wtf gene can be detoxified by the original wtf gene. Like newly duplicated genes, the most probable fate of a new wtf duplicate is pseudogenization (Lynch, 2007; Innan and Kondrashov, 2010). Thus, the vast majority of new wtf duplicates experience an early exit from the population, most probably never reaching fixation (Lynch, 2007; Innan and Kondrashov, 2010). To become a new driver, the new wtf copy should evolve coupling new poison and new antidote to the new poison through mutations. But the fate-changing mutations are likely to be rare. It follows that gene duplication might be insufficient to drive the diversification of wtf genes.

In this study, we analyzed the diversity and evolution of wtf genes in the genomes of 21 strains of S. pombe that were sequenced using long-read sequencing approaches (Tusso et al., 2022). Through knocking out each of 25 wtf genes in S. pombe laboratory strain 972h-, no significant attenuated growth was observed, indicating wtf genes might be not deleterious in the asexual life cycle. We found that recurrent recombination occurred among wtf genes. We generated a novel meiotic driver through artificial recombination between wtf drivers, and its encoded poison cannot be detoxified by the antidotes encoded by their parental wtf genes but can be detoxified by its own antidote. Therefore, we propose that recombination can generate wtf driver with new poisons and might shape the diversification of wtf genes.

Results

Diversity and evolution of wtf genes in fission yeasts

First, we analyzed the diversity and distribution of wtf genes in fission yeasts. The S. pombe reference genome (strain 972h-) encodes a total of 25 wtf genes. For these 25 wtf genes, the number of exons varies from 3 to 6 (Figure 1A) (Bowen et al., 2003). To investigate the relationship among exons from different wtf genes, we grouped these wtf exons into clusters based on nucleotide identity of 0.50 (Figure 1B and C). The wtf exons were grouped into 10 clusters with >2 members, and 12 exons exist as singletons in the similarity network (Figure 1B and C; Supplementary file 1b). Only exon 1 of all the 25 wtf genes group together in a cluster, indicating that the first exons are well conserved among the wtf genes. No other exon is conserved among all the 25 wtf genes. Therefore, the evolution of the wtf gene structures appear to be highly dynamic.

The diversity and evolution of wtf genes in fission yeasts.

(A) The gene structures of the wtf genes in the S. pombe reference genome. Rectangles and lines represent exons and introns, respectively. Rectangles with the same color indicate exons that share >50% sequence identity. (B) The similarity network of exons of the wtf genes in the S. pombe reference genome. Sequences that share an identity of >50% form a cluster. The colors of wtf exons correspond to these in panel (A). The exons for each cluster were listed in the Supplementary file 1b. (C) The heatmap of nucleotide identity among exons of the wtf genes in S. pombe reference genome. (D) The distribution wtf genes in 21 S. pombe strains. wtf genes were present in 25 genetic loci. Filled circles and empty circles represent the presence or absence of wtf genes, respectively. Red pentagons indicate putative wtf pseudogenes. The size of circles in purple indicates the number of wtf genes. The relationship among 21 S. pombe strains was inferred based on phylogenetic analysis of 30 randomly selected genes. (E) Phylogenetic relationship of wtf genes in 21 S. pombe strains. The phylogenetic tree is reconstructed using the maximum likelihood method. Filled circles in blue indicate wtf genes in S. pombe reference genome. The number of exons, the presence of solo-LTRs, the putative pseudogene status, the presence of intron-1-ATG codon, the presence of exon-2-ATG codon are shown near the corresponding wtf gene. wtf genes from other fission yeast species were collapsed into a triangle.

We next identified wtf genes in 21 strains of S. pombe that were sequenced using long-read sequencing approaches (Supplementary file 1c) (Tusso et al., 2022). The copy number of wtf genes varies among different S. pombe strains, ranging from 24 (strain JB879) to 37 (strain JB1206) (Figure 1D). Synteny analyses show that the wtf genes are present in 20 genetic loci (Figure 1D). Multiple wtf genes were present in 13 wtf loci. Within 20 wtf loci, at least one wtf gene is present in all of or nearly all of the 21 S. pombe strains, suggesting that these 20 wtf loci might have originated before the LCA of the 21 S. pombe strains. wtf pseudogenes are prevalent in many wtf loci among the 21 S. pombe strains, indicating frequent pseudogenization occurred in the wtf genes. These results indicate that wtf copy number variation is prevalent among S. pombe strains.

We performed phylogenetic analyses of the wtf genes from 21 S. pombe strains and three other fission yeast species (S. octosporus, S. cryophilus, and S. osmophilus) (Figure 1E; Supplementary file 1d). The wtf genes of S. pombe form a monophyletic group. Orthologs of wtf14, wtf7, wtf11, and wtf15 form monophyletic groups, whereas orthologs of other wtf genes show complex phylogenetic mixing, indicating complex recombination might have occurred among these wtf genes (Figure 1E) (Eickbush et al., 2019). Moreover, the wtf genes with 6 exons (including the known meiotic drivers wtf4, wtf9, wtf13 and wtf23) (Nuckolls et al., 2017; Bravo Nunez et al., 2018a; Bravo Nunez et al., 2020a) cluster together, and exhibit a ladder-like phylogeny, which might be generated by continual selection driven by antidotes (like the ladder-like phylogeny of influenza A viruses H1N1 and H3N2, which is shaped by continual immune selection (Grenfell et al., 2004; Bedford et al., 2011)). Based on phylogenetic relationship, we divided the wtf genes of 21 S. pombe strains into eight groups, namely groups 1 to 8, among which groups 5 to 8 include orthologs of wtf14, wtf7, wtf11, and wtf15, respectively (Figure 1E). Exon 2 ATG codons (exon-2-ATG) and in-frame ATG within intron 1 and near the start of exon 2 (intron-1-ATG) of wtf genes can encode the start of poison protein isoforms (Hu et al., 2017). We found that most of exon-2-ATG and intron-1-ATG are present within group 1 wtf genes (Figure 1E). A majority of the wtf genes are flanked by solo-LTRs (Bowen et al., 2003) (Figure 1E). However, the solo-LTRs flanking the wtf genes do not cluster together but form many distinct groups, suggesting that solo-LTRs were inserted nearby the wtf genes multiple times (Supplementary file 1a). Together, our results reveal the rapid diversification and turnover of wtf genes in a single fission yeast species.

No attenuated growth of fission yeast without wtf genes

To explore the effect of wtf genes on the fitness of fission yeast, we knocked out each of the 25 wtf genes in the S. pombe laboratory strain 972h-using a method based on homologous recombination (Figure 2A). A total of 25 wtf knockout strains (Δwtf1 to Δwtf25) were generated. We used spot assay to evaluate the effect of wtf gene knockout on the yeast growth, and no growth defect was observed for all the 25 wtf knockout strains (Figure 2B). Furthermore, no significant differences were observed in the growth curves between the wild-type and wtf knockout strains (Figure 2C) or in the maximum growth rates among the wild-type and wtf knockout strains. Therefore, our experiment suggests that the wtf genes might be largely neutral to the fitness of their carriers in the asexual life cycle at least in normal growth condition.

The effect of wtf gene on the growth of S. pombe.

(A) Generation of wtf knockout (Δwtf) strains based on the homologous recombination method. Substitution cassette contains a kanMX resistance marker and two homologous sequences flanking the target wtf genes. (B) Spot assay of Δwtf strains. The strains were diluted in five 10-fold steps to 10−5, and 1.5μL of each dilution were spotted on the surface of YE solid media. Growth curves (C) and maximum growth rates (D) of WT and 25 wtf knockout strains. Data represent means of three biological replicates (solid lines or bars), with error bars showing SD. Open circles indicate individual replicate values.

Recurrent recombination in wtf genes

Given complex phylogenetic mixing observed among wtf genes (Figure 1E), we tested whether recombination occurred. We detected signals of recombination in the 25 wtf genes of the S. pombe reference genome (p <0.0001) and in the wtf genes of the 21 S. pombe strains (p <0.0001) using pairwise homoplasy index (HPI) test. Split network analysis also supports the frequent occurrence of recombination in the 25 wtf genes of the S. pombe reference genome (Figure 3A) and in the wtf genes of the 21 strains of S. pombe (Figure 3B). In contrast, no recombination signal was detected for groups 5 to 8 using HPI test (p = 1 for group 5, p = 1 for group 6, p = 0.53 for group 7, and p = 1 for group 8). We estimated recombination rates of the full-length wtf sequences, the first exons, and the wtf sequences without the first exons for wtf groups 1 to 4. We found that the recombination rate of group 1 wtf was highest among the four wtf groups (Figure 3C). For group 1, breakpoints are dispersed across the wtf sequences (Figure 3D). These lines of evidence suggest that wtf genes underwent recurrent and intricate recombination.

Recombination analysis of wtf genes.

(A) Split tree of 25 wtf genes of the S. pombe reference genome. (B) Split tree of wtf genes from 21 S. pombe strains. The wtf genes of the S. pombe reference genome are labelled. (C) Recombination rates of wtf genes that belong to groups 1 to 4. Recombination rates were estimated for the full-length wtf sequences, the first exons, and the wtf sequences without the first exons. (D) Breakpoints detected for wtf genes of group1. The exons of wtf4, as a gene position reference, are shown.

Generation of a new driver gene through artificial recombination

Given gene duplication alone might be insufficient to shape the diversification of wtf genes, we hypothesize that recombination between wtf genes can generate new meiotic drives. To test this, we constructed four chimeric wtf genes through recombination among known functional meiotic drivers (wtf23) (Bravo Nunez et al., 2018a; Bravo Nunez et al., 2020a) and an artificially generated meiotic driver (wtf18) as specified below.

We used a proved Saccharomyces cerevisiae system to test the activity of poison and antidote proteins encoded by wtf genes (Nuckolls et al., 2020). As expected, the expression of the poison proteins (Wtf23poison) encoded by wtf23 genes caused the yeast growth arrest (Figure 4A). The attenuated growth was alleviated, when the corresponding antidote proteins (Wtf23antidote) were expressed (Figure 4A). We also experimentally analyzed wtf18 gene, which was known to encode only long (antidote-like) transcripts and probably act as a suppressor (Bravo Nunez et al., 2018a). We artificially introduced an in-frame ATG codon right before the start of exon 2, generating wtf18poison/-0M. The expression of wtf18poison/-0M resulted in the yeast growth arrest, suggesting its product, Wtf18poison/-0M, is indeed a poison protein (Figure 4B). When co-expressing wtf18antidote and wtf18poison/-0M, the attenuated yeast growth was rescued (Figure 4B), indicating that Wtf18antidote can ameliorate the toxicity of Wtf18poison/-0M.

Poison and antidote activity of wtf genes and chimeric wtf genes.

wtf23 and wtf18 are highlighted in green and red, respectively. Rectangles represent exons. The start codon ATG is shown. For wtf18, an in-frame ATG codon was introduced right before the start of exon 2, generating wtf18poison/-0M. Spot assay of yeast transformed with the short isoforms (encoding poison-like proteins) and the long isoforms (encoding antidote-like proteins) of wtf23 (A) and wtf18 (B). (C to F) Four chimeric wtf genes were generated through artificial recombination, namely wtfC1 (with exons 1-2 of wtf23 and exons 3-6 of wtf18), wtfC2 (with exons 1-3 of wtf23 and exons 4-6 of wtf18), wtfC3 (with exons 1-4 of wtf23 and exons 5-6 of wtf18), and wtfC4 (with exons 1-5 of wtf23 and exon 6 of wtf18). Spot assay of yeast transformed with the short isoforms (encoding poison-like proteins) and the long isoforms (encoding antidote-like proteins) of wtfC1 (C), wtfC2 (D), wtfC3 (E), and wtfC4 (F) are shown.

We then constructed four chimeric wtf genes through artificial recombination between wtf23 and wtf18, including wtfC1 (possessing exons 1-2 of wtf23 and exons 3-6 of wtf18), wtfC2 (possessing exons 1-3 of wtf23 and exons 4-6 of wtf18), wtfC3 (possessing exons 1-4 of wtf23 and exons 5-6 of wtf18), and wtfC4 (possessing exons 1-5 of wtf23 and exon 6 of wtf18). The expression of the short isoforms of wtfC1, wtfC2, wtfC3, and wtfC4 resulted in yeast growth arrest, revealing their toxicity (Figure 4C-F). However, the antidote of wtfC1 and wtfC3 cannot detoxify the corresponding chimeric toxins (Figure 4C and E). Interestingly, we generated a putative novel meiotic driver, namely wtfC4. Our results show that wtfC4 encodes a functional poison (WtfC4poison) (Figure 4F). The poison can be detoxified by its own long isoforms (dubbed as wtfC4antidote), but cannot be detoxified by the antidote proteins of their parental genes (Figure 4F). Taken together, we generated a new meiotic driver through artificial recombination between pre-existing wtf genes.

We tried to test the driver phenotype of wtfC4 in a more natural setting. We created a recombinant strain, Sp-wtfC4, based on the laboratory strain 972h-. Specifically, we replaced the last exon of the original wtf23 gene with the last exon of wtf18. However, we encountered a challenge: since strain 972h-has only one mating type and cannot undergo meiosis on its own, we had to mate the recombinant strain with a BN0 h+ strain that only carries wtf23antidote. We did not observe meiotic driver phenotype as expected. This might be due to issues with the proper splicing and expression of the potential poison and antidote proteins or due to the genetic background. Nevertheless, our results raise the possibility that new meiotic drivers can arise through recombination.

Discussion

In this study, we analyzed the diversity and evolution of wtf genes in fission yeasts. The copy number of the wtf gene varies among different S. pombe strains, revealing rapid diversification and turnover of the wtf genes within a single fission yeast species. We detected signals of recurrent and intricate recombination among wtf genes as previously reports with limited genomes (Hu et al., 2017; Nuckolls et al., 2017; Eickbush et al., 2019; De Carvalho et al., 2022). We hypothesize that recombination between wtf genes can produce new wtf genes with new poisons and the antidotes to new poisons. These new wtf genes can then driver through populations. As proof-of-principle, we generated a chimeric wtf gene that represents a new meiotic driver. The encoded poison of the newly generated meiotic driver can be detoxified by its own long isoforms, but cannot be detoxified by the antidote proteins of their parental genes. However, the other three chimeric wtf genes tested did not show this property. Indeed, our recombination breakpoint analyses (Figure 3D) reveal substantial recombination might have occurred in the last exon. Together, our results indicate that recombination is likely to drive the rapid diversification of wtf gene in fission yeasts.

Most of the known meiotic drivers impose costs on their carriers due to direct effects of the driver on survival or fertility, production of a biased sex ratio, or via deleterious mutations linked to the driver (Price and Wedell, 2008; Larracuente et al., 2012; Sutter and Lindholm, 2015; Fishman et al., 2015; Lindholm et al., 2016; Zanders and Unckless, 2019). In outcrossing between individuals from distinct yeast lineages, wtf drivers can provide a selective advantage to atypical spores, such as aneuploids and diploids (Bravo Nunez et al., 2020b). In this study, we assessed the effects of the wtf genes on the growth of fission yeast during the asexual life cycle through knocking out each of the 25 wtf genes in S. pombe laboratory strain 972h-separately. We did not observe obvious attenuated growth for these wtf knockout strains, indicating wtf genes are largely neutral to the fitness of their carriers during the asexual life cycle at least in the normal growth setting. It should be noted that the spot assay used in this study detects only large differences in fitness between wide type and wtf knockout strains. Nevertheless, it is likely that wtf genes evolve mainly in a neutral manner during the asexual life cycle, which explains the presence of high proportion of pseudogenes in wtf gene repertoire. Moreover, asexual reproduction is much more frequent than sexual reproduction for yeasts (Tsai et al., 2005). Therefore, even if fate-changing mutations that simultaneously produce new poison and the antidote to new poison occur, the most probable fate of a new wtf gene generated by gene duplication is pseudogenization and removal from the population.

Gene duplication gives rise to new wtf genes. However, the newly generated wtf gene can be detoxified by the original wtf gene and thus cannot drive through its host population, when the original wtf is fixed in the population (Figure 5). Therefore, most, if not all, of wtf gene duplicates experience early exist from the host population. When recombination occurs between two pre-existing wtf genes, chimeric wtf gene with new poison and the antidote to new poison can be generated as this study shows. Then, the wtf gene with new driver property can spread in its host population, even reaching fixation (Figure 5). During asexual life cycle, wtf genes evolve mainly under genetic drift, and thus can accumulate disruptive mutations, leading to their pseudogenization. Taken together, our study highlights the significance of recombination in shaping the diversification of wtf genes.

The evolutionary mechanisms of wtf genes.

(A) Mechanism of wtf meiotic driver. wtf genes are transcribed into two transcripts, namely long and short isoforms. The long and short isoforms encode antidote and poison, respectively. All spores are exposed to the poison, whereas only those that inherit wtf genes express the antidote and are rescued. Spores without the wtf allele are destroyed, resulting in the drive of wtf. (B) Duplication is insufficient to drive the diversification of wtf genes. Gene duplication of wtfA gives rise to a new wtf gene, wtfAdup. However, wtfAdup gene can be detoxified by the original wtfA gene and thus cannot drive through its host population. (C) When recombination occurs between two parental wtfB and wtfC genes, generating chimeric wtfD gene. wtfD encodes new poison that cannot be detoxified by their parental antidote and the antidote to new poison. The wtfD gene then can spread in its host population.

Methods

Identification of the wtf genes

We used the blastn algorithm to identify wtf genes within 21 S. pombe strains with 25 wtf genes from S. pombe reference genome as the queries and an e-cutoff value of 10−5. The identified wtf genes were annotated based on the wtf genes of reference genome. The sequence identity among the exons of wtf genes was calculated using BioAider version 1.334 (Zhou et al., 2020). Exons was then clustered based on the nucleotide identity using igraph package version 2.0.1.1 (Csardi et al., 2006; Csardi et al., 2024). We extended 1000 bp flanking each wtf gene to establish their syntenic relationships.

Phylogenetic analysis

The coding sequences of wtf genes of 21 S. pombe strains and three other fission yeast species (S. octosporus, S. cryophilus, and S. osmophilus) were aligned using MAFFT version 7 (Katoh and Standley, 2013). To clarify the relationship of 21 S. pombe strains, 30 genes were randomly selected and concatenated using Phylosuite version 1.2.1 (Zhang et al., 2020). All the phylogenetic analyses in this study were performed using the maximum likelihood (ML) method implemented in IQ-TREE version 2 (Minh et al., 2020). The best-fit substitution model was selected using the ModelFinder algorithm (Kalyaanamoorthy et al., 2017). Node supports were assessed using the ultrafast bootstrap approximation (UFBoot) method with 1,000 replicates (Hoang et al., 2018). Solo-LTRs were identified using the blast algorithm and aligned using MAFFT version 7 (Katoh and Standley, 2013). Phylogenetic analysis was performed using the approximate maximum likelihood method implemented in FastTree version 2.1.1 (Price et al., 2010).

Recombination analysis

Split networks of wtf genes were generated using the neighborhood network analysis implemented in SplitsTree4 (Huson et al., 2006). Pairwise homoplasy index (PHI) test was performed using SplitsTree4 (Huson et al., 2006). Potential breakpoints were detected using 3SEQ (Lam et al., 2018). The recombination rate was estimated using FastEPRR package version 2.0 (Gao et al., 2016). Recombination rate was estimated using FastEPRR version 2.0 (Gao et al., 2016) as the population-scaled recombination rate, Rho = 4Ner, where Ne is the effective population size and r is the per-generation recombination rate. This scaling allows comparison of recombination rates across genomic regions and populations.

Generation of wtf knockout strains

The wtf knockout (Δwtf) strains generated in this study were derived from S. pombe strain 972h-. We constructed substitution cassettes for each of the 25 wtf genes of the S. pombe reference genome. Substitution cassettes contain a kanMX resistance marker and two homologous sequences flanking the target wtf genes (Moreno et al., 1991; García-Ríos et al., 2014). Substitution cassettes were transformation into S. pombe (strain 972h-) through the lithium acetate-based method (Moreno et al., 1991; García-Ríos et al., 2014). wtf gene knockout strains were selected for kanMX resistance and were verified by PCR. The primers used for wtf knockout in S. pombe are provided in Supplementary file 1f.

Plasmid construction

Total RNA of fission yeast was extracted and reverse transcribed into cDNA. Coding sequences of wtf23antidote, wtf23poison, and wtf18antidote were amplified using the corresponding primers (Supplementary file 1g). wtf18poison/-M0 was generated using wtf18antidote as the template and the primer with an artificially introduced ATG (Supplementary file 1g). We generated wtfC1 through recombining exons 1-2 of wtf23 and exons 3-6 of wtf18, generated wtfC2 through recombining exons 1-3 of wtf23 and exons 4-6 of wtf18, generated wtfC3 through recombining exons 1-4 of wtf23 and exons 5-6 of wtf18, and generated wtfC4 through recombining exons 1-5 of wtf23 and exon 6 of wtf18. These wtf and wtfC genes were then cloned into the GAL1/10 dual expression plasmid Gal_HF. Plasmids were first transformed into Escherichia coli and verified by PCR and sequencing. Plasmids were then transformed into S. cerevisiae (strain S288C) using the lithium acetate-based method. Yeast transformants were selected for kanMX resistance and were verified by PCR.

Meiotic analysis

We created a recombinant strain, Sp-wtfC4, based on the laboratory strain 972h-. Specifically, we replaced the last exon of the original wtf23 gene with the last exon of wtf18 using homologous recombination. The SP-wtfC4 strain and the BN0 h+ strain carrying wtf23antidote were streaked separately onto YE solid plates and incubated at 30°C for approximately 20 hours. Cells from each strain were then scraped and resuspended in sterile water to an OD600 of approximately 0.5, and mixed in equal volumes. The mixture was spread onto SPA plates and incubated upside down at 30°C for 2–3 days to induce meiosis and sporulation. For the fertility assay, 5–10 μL of propidium iodide (PI, 1 mg/mL) was added to 50 μL of H2O, and cells were scraped from the SPA plates and suspended in the PI mix. The mixture was incubated at room temperature for 30 minutes, followed by gentle centrifugation to collect the cells. Fluorescence microscopy was then used for observation and imaging (Nuckolls et al., 2017).

Spot assay

The yeast strains were cultured in YPD liquid medium at 30°C with shaking at 200 rpm. The overnight cultures were transferred to fresh YPD liquid medium and grown to an OD600 value of ~3. Cells were collected by centrifugation, and the OD600 was adjusted to 3. Subsequently, the strains were diluted in five 10-fold steps to 10−5, and 1.5μL of each dilution were spotted on the surface of YPD and YPG solid media. The plates were incubated at 30°C, and the growth of colonies were observed.